---
title: "DDS Case Study 2"
author: "Ganesh, Jeff and Stacey"
date: "November 20, 2018"
output:
  html_document: default
  word_document: default
---

### Background
DDSAnalytics is an analytics company that specializes in talent management solutions for Fortune 1000 companies. Talent management is defined as the iterative process of developing and retaining employees. It may include workforce planning, employee training programs, identifying high-potential employees and reducing/preventing voluntary employee turnover (attrition). To gain a competitive edge over its competition, DDSAnalytics is planning to leverage data science for talent management. The executive leadership has identified predicting employee turnover as its first application of data science for talent management.They wanted to conduct an analysis of existing employee data to contain the employee turnover
DDSAnalytics approached fine data science team from SMU including Jeff Leath, Stacey Smith and Ganesh Kodi to conduct exploratory data analysis to determine factors that lead to attrition using existing employee dataset
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

## Load packages
library(plyr)
library(tidyr)
library(dplyr)
library(base)
library(ggplot2)
library(gridExtra)
library(sqldf)


```

### Code 
#### Clean the Raw data 
Convert source data provided by DDAnalytics and convert it to csv. 
Read the csv file and create dataframe from csv

Following are the structure of the dataset. 
Noticed that number of rows are 1470 and columns are 35

```{r ddanalytics summary, echo=TRUE}
## Read the attrition data provided by DDSAnalytics into dataframe
df_ddanalytics = read.csv("Source/CaseStudy2-data.csv", header = TRUE)

## Dimension of the dataframe which provides number of rows and columns
dim(df_ddanalytics)

```

Restrict the lengthy column name within 12 characters. Updated column names are 
below. We will replace lookup columns (like JobSatisfct,EnvSatisfct etc) values with the actual description provided in the dataset. Removed non significant columns and create new dataset for Exploratory Analysis

```{r tidying ddanalytics, echo=TRUE}
## Change column names of the dataframe within 12 characters
names(df_ddanalytics) <-c("Age","Attrition","BusinsTravl","DailyRate","Department","DistancHome","Education","EducField","EmployCount","EmployeeNum","EnvSatisfct","Gender","HourlyRate","JobInvolve","JobLevel","JobRole", "JobSatisfct","MaritlStats","MonthIncom","MonthRate", "NumCompWorkd","Over18", "OverTime","SalaryHike","PerfRating","RelSatisfct","StandHours","StockOption","WorkingYear","LastYrTrain","WorkLifeBal","YrsAtCompny","CurRoleYrs","YrslastProm","YrsCurMangr")
## Convert the parameters to descrption shared in definitions and create new dataset
df_ddanalytics_updated <- df_ddanalytics %>%
  mutate(Education = as.factor(if_else(Education == 1,"Below College", if_else(Education == 2, "College", if_else(Education == 3, "Bachelor", if_else(Education == 4, "Master","Doctor")))))
         ,EnvSatisfct = as.factor(if_else(EnvSatisfct == 1,"Low",if_else(EnvSatisfct == 2, "Medium", if_else(EnvSatisfct == 3, "High", "Very High"))))
         ,JobInvolve = as.factor(if_else(JobInvolve == 1,"Low",if_else(JobInvolve == 2, "Medium",if_else(JobInvolve == 3, "High", "Very High"))))
         ,JobSatisfct = as.factor(if_else(JobSatisfct == 1, "Low",if_else(JobSatisfct == 2, "Medium",if_else(JobSatisfct == 3, "High","Very High"))))
         ,PerfRating = as.factor(if_else(PerfRating == 1, "Low",if_else(PerfRating == 2, "Good", if_else(PerfRating == 3, "Excellent", "Outstanding"))))
         ,RelSatisfct = as.factor(if_else(RelSatisfct == 1, "Low",if_else(RelSatisfct == 2, "Medium", if_else(RelSatisfct == 3, "High", "Very High"))))
         ,WorkLifeBal = as.factor(if_else(WorkLifeBal == 1, "Bad",if_else(WorkLifeBal == 2, "Good", if_else(WorkLifeBal == 3, "Better", "Best"))))
   )%>% select(-EmployCount, -EmployeeNum, -Over18, -StandHours, -StockOption, -JobLevel)
summary(df_ddanalytics_updated)
```

Above data provides high level view of dataset. Attrition rate of DDAnalytics with current data is 237/1470 i.e 16.1%

#### Preliminary Analysis
As part of the preliminary analysis, found that dataset doesn't have under age (<18 years age) data. Also, noticed that maximum age in the dataset is of 60 years. Considering this as permitted age and disregard the notion of age outlier in the dataset.
```{r ddanalytics prelim analysis, echo=TRUE}
max(df_ddanalytics$Age)
min(df_ddanalytics$Age)
```

Identified the 7 important explanatory variables with (Age, MonthlyIncome, JobSatisfaction, DailyRate, PercentSalaryHike, WorkLifeBalance, YearsInCurrentRole, YearsSinceLastPromotion, TotalWorkingYears, RelationshipSatisfaction) with response variable as Attrition. 
Created seperate datset based on these important varaibles.  Descriptive Statistics of these variables provided below.

```{r ddanalytics descriptive total, echo=TRUE}
## Create sub data frame of limited variables
dfanalytics <- data.frame(df_ddanalytics$Age,df_ddanalytics$MonthIncom,df_ddanalytics$JobSatisfct,df_ddanalytics$DailyRate,df_ddanalytics$SalaryHike,df_ddanalytics$WorkLifeBal,df_ddanalytics$CurRoleYrs,df_ddanalytics$YrslastProm,df_ddanalytics$WorkingYear,df_ddanalytics$RelSatisfct, df_ddanalytics$Attrition)
## Adding column names for the sub data frame
names(dfanalytics) <- c("Age","MonthIncom","JobSatisfct","DailyRate","SalaryHike","WorkLifeBal","CurRoleYrs","YrslastProm","WorkingYear","RelSatisfct","Attrition")
dfanalytics_attr <- subset(dfanalytics, Attrition == "Yes")
dfanalytics <- dfanalytics[-c(11)]
dfanalytics_attr <- dfanalytics_attr[-c(11)]
## Create table of statistics of all
stattable <- do.call(data.frame, 
           list(mean = apply(dfanalytics, 2, mean),
                standdev = apply(dfanalytics, 2, sd),
                median = apply(dfanalytics, 2, median),
                min = apply(dfanalytics, 2, min),
                max = apply(dfanalytics, 2, max),
                length = apply(dfanalytics, 2, length)))
## Display the statistics of each variables
stattable
```

We have taken Age and Income for plotting histogram. Plot shows that Age distribution is somewhat normal though there is little bit of right skewed. But, More right skewed grapgh of Monthly Income of the employees shows that non normal distribution. Also there is presence of outlier in the range of 20000. Monthly Inbcome should be transformed to curb the skewness.

```{r histogram ddanalytics, echo=TRUE}
## Histogram of Age
hist(dfanalytics$Age, xlab = "Age", main = "Histogram of Age", col="lightgreen")
## Histogram of Monthly Income
hist(dfanalytics$MonthIncom, xlab = "Monthly Income", main = "Histogram of Monthly Income", col ="lightblue" )

```

Frequencies of the Gender, Education and Occupation are below.
```{r Frequency DDAnalytics, echo=TRUE}
## Frequency of the Gender
df_ddanalytics_updated %>% group_by(Gender) %>%  summarise(Freq = n()) 
## Frequency of the Education
df_ddanalytics_updated %>% group_by(Education) %>%  summarise(Freq = n()) 
## Frequency of the Occupation
df_ddanalytics_updated %>% group_by(JobRole) %>%  summarise(Freq = n()) 

```

Counts of Management positions are listed below. We assumed that Management positions consists of Manager / Director Job Role. 
```{r Management Count DDAnalytics Attrition, echo=TRUE}
df_ddanalytics_updated %>% filter(JobRole == "Manager" | JobRole == "Research Director" | JobRole == "Manufacturing Director"  ) %>% group_by(JobRole) %>%  summarise(Freq = n()) 
```

#### Deeper Analysis and Visualization
As part of the deeper analysis, we will analyse each variable independently by creating bar charts. Bar charts will be displayed in descending order
```{r ddanalytics updated deeper analysis, echo=TRUE, warning=FALSE}
grid.arrange(
df_ddanalytics_updated %>% group_by(Age) %>% summarise(counts = n()) %>% ggplot(aes(x = factor(Age,levels = Age[order(-counts)]), y = counts)) +xlab("Age") + geom_histogram(stat = 'identity', binwidth = 5,fill = "violet")+ theme(axis.text.x = element_text(angle = 90, hjust = 1)),

df_ddanalytics_updated %>% group_by(DistancHome) %>% summarise(counts = n()) %>%ggplot(aes(x = factor(DistancHome,levels = DistancHome[order(-counts)]), y = counts)) +xlab("Distance From Home") + geom_histogram(stat = 'identity', binwidth = 5,fill = "lightblue")+ theme(axis.text.x = element_text(angle = 90, hjust = 1)),
 ncol = 1, nrow = 2)

grid.arrange(
df_ddanalytics_updated %>% group_by(NumCompWorkd) %>% summarise(counts = n()) %>% ggplot(aes(x = factor(NumCompWorkd,levels = NumCompWorkd[order(-counts)]), y = counts)) +xlab("Number of companies worked") + geom_histogram(stat = 'identity', binwidth = 2,fill = "orange")+ theme(axis.text.x = element_text(angle = 90, hjust = 1)),

df_ddanalytics_updated %>% group_by(WorkingYear) %>% summarise(counts = n()) %>%ggplot(aes(x = factor(WorkingYear,levels = WorkingYear[order(-counts)]), y = counts)) +xlab("Total Working Years") + geom_histogram(stat = 'identity', binwidth = 4,fill = "lightgreen")+ theme(axis.text.x = element_text(angle = 90, hjust = 1)),  
 ncol = 1, nrow = 2)
grid.arrange(
      ggplot(df_ddanalytics_updated) + geom_histogram(aes(MonthIncom), binwidth = 1000, fill = "violet",col = "black")+xlab("Monthly Income") +theme(axis.text.x = element_text(angle = 90, hjust = 1)),
    
    df_ddanalytics_updated %>% group_by(YrsAtCompny) %>% summarise(counts = n()) %>%ggplot(aes(x = factor(YrsAtCompny,levels = YrsAtCompny[order(-counts)]), y = counts)) +xlab("Total Working Years") + geom_histogram(stat = 'identity', fill = "orange")+ theme(axis.text.x = element_text(angle = 90, hjust = 1)),
ncol = 1, nrow = 2)
grid.arrange(
  df_ddanalytics_updated %>% group_by(CurRoleYrs) %>% summarise(counts = n()) %>% ggplot(aes(x = factor(CurRoleYrs,levels = CurRoleYrs[order(-counts)]), y = counts)) +xlab("Years in Current Role") + geom_histogram(stat = 'identity', fill = "lightgreen")+ theme(axis.text.x = element_text(angle = 90, hjust = 1)),
    
    df_ddanalytics_updated %>% group_by(YrslastProm) %>% summarise(counts = n()) %>%ggplot(aes(x = factor(YrslastProm,levels = YrslastProm[order(-counts)]), y = counts)) +xlab("Years Since Last Promotion") + geom_histogram(stat = 'identity', fill = "lightblue")+ theme(axis.text.x = element_text(angle = 90, hjust = 1)),

ncol = 1, nrow = 2)
grid.arrange(
  df_ddanalytics_updated %>% group_by(YrsCurMangr) %>% summarise(counts = n()) %>% ggplot(aes(x = factor(YrsCurMangr,levels = YrsCurMangr[order(-counts)]), y = counts)) +xlab("Years with current manager") + geom_histogram(stat = 'identity', fill = "brown")+ theme(axis.text.x = element_text(angle = 90, hjust = 1)),
    
    df_ddanalytics_updated %>% group_by(SalaryHike) %>% summarise(counts = n()) %>%ggplot(aes(x = factor(SalaryHike,levels = SalaryHike[order(-counts)]), y = counts)) +xlab("Percent Salary Hike") + geom_histogram(stat = 'identity', fill = "yellow")+ theme(axis.text.x = element_text(angle = 90, hjust = 1)),

 ncol = 1, nrow = 2)

grid.arrange(
  df_ddanalytics_updated %>% group_by(Education) %>% summarise(counts = n()) %>% ggplot(aes(x = factor(Education,levels = Education[order(-counts)]), y = counts)) +xlab("Education") + geom_histogram(stat = 'identity', fill = "violet")+ theme(axis.text.x = element_text(angle = 90, hjust = 1)),
    
    df_ddanalytics_updated %>% group_by(EducField) %>% summarise(counts = n()) %>%ggplot(aes(x = factor(EducField,levels = EducField[order(-counts)]), y = counts)) +xlab("Education Field") + geom_histogram(stat = 'identity', fill = "lightblue")+ theme(axis.text.x = element_text(angle = 90, hjust = 1)),
  
  df_ddanalytics_updated %>% group_by(RelSatisfct) %>% summarise(counts = n()) %>% ggplot(aes(x = factor(RelSatisfct,levels = RelSatisfct[order(-counts)]), y = counts)) +xlab("Relationship Satisfaction") + geom_histogram(stat = 'identity', fill = "lightgreen")+ theme(axis.text.x = element_text(angle = 90, hjust = 1)),
    
    df_ddanalytics_updated %>% group_by(WorkLifeBal) %>% summarise(counts = n()) %>%ggplot(aes(x = factor(WorkLifeBal,levels = WorkLifeBal[order(-counts)]), y = counts)) +xlab("Work life Balance") + geom_histogram(stat = 'identity', fill = "orange")+ theme(axis.text.x = element_text(angle = 90, hjust = 1)),
  
  df_ddanalytics_updated %>% group_by(BusinsTravl) %>% summarise(counts = n()) %>% ggplot(aes(x = factor(BusinsTravl,levels = BusinsTravl[order(-counts)]), y = counts)) +xlab("Business Travel") + geom_histogram(stat = 'identity', fill = "red")+ theme(axis.text.x = element_text(angle = 90, hjust = 1)),
    
    df_ddanalytics_updated %>% group_by(EnvSatisfct) %>% summarise(counts = n()) %>%ggplot(aes(x = factor(EnvSatisfct,levels = EnvSatisfct[order(-counts)]), y = counts)) +xlab("Environment Satisfaction") + geom_histogram(stat = 'identity', fill = "yellow")+ theme(axis.text.x = element_text(angle = 90, hjust = 1)),

 ncol = 3, nrow = 2)
```

As part of the deeper analysis, we will check the relationship between age and income

```{r ddanalytics age and monthly histogram, echo=TRUE}

## Scatterplot between Age and Income 
ggplot(df_ddanalytics_updated, aes(Age,MonthIncom,color=Gender))+geom_point()+ggtitle("Relation of Age vs Monthly Income")+labs(x="Age", y="Income")

```
Above scatter plot between Age and Monthly Income shows that there is positive linear corelation between Age and Income. 
Below linear regression summary shows that there is significance relation between Age and Income. Gender category doesn't have any signifiance in the model

```{r ddanalytics age income summary, echo=TRUE}
## Summary of linear regression model of Age and Income and combined with Gender
dd_ageincome.lm <- lm(Age~MonthIncom*Gender, data=df_ddanalytics_updated)
summary(dd_ageincome.lm)

```

```{r ddanalytics age vs job satisfaction, echo=TRUE}
## Scatterplot between Age and Job Satisfaction 
ggplot(df_ddanalytics_updated, aes(Age,JobSatisfct,color=Gender))+geom_point()+ggtitle("Relation of Age vs Job Satisfaction")+labs(x="Age", y="Job Satisfaction")

```

Above scatter plot between Age and Job Satisfaction shows that there is no linear corelation between Age and Job Satisfaction 
Below linear regression summary shows that there is no significance relation between Age and Job Satisfaction. Gender category doesn't have any signifiance in the model

```{r ddanalytics age jobsatisafction summary, echo=TRUE}
## Summary of linear regression model of Age and Jobsatisfaction and combined with Gender
dd_agejobsatisfact.lm <- lm(Age~JobSatisfct*Gender, data=df_ddanalytics_updated)
summary(dd_agejobsatisfact.lm)

```
## The Plots Show that attrition employees were likely to be single and worked more overtime.
```{r compare Marital Status and Overtime with Attrition and Non Attition Employees, echo=TRUE}

df_ddanalytics = read.csv("Source/CaseStudy2-data.csv", header = TRUE)
attrition_Employees <- sqldf("select * from df_ddanalytics where Attrition = 'Yes'")
no_attrition_Employees <- sqldf("select * from df_ddanalytics where Attrition = 'No'")

df_summary1 <-
  attrition_Employees                          %>% # Pipe df into group_by
  group_by(MaritalStatus)              %>% # grouping by 'type' column
  summarise(name_count = n())     # calculate the name count for each group

df_summary2 <-
  no_attrition_Employees                          %>% # Pipe df into group_by
  group_by(MaritalStatus)              %>% # grouping by 'type' column
  summarise(name_count = n())     # calculate the name count for each group


## (1) Plot pre summarized data: 'df_summary'.
plot1 <- ggplot(df_summary1, aes(MaritalStatus, name_count, fill=MaritalStatus)) +  #
  geom_bar(stat = 'identity')+              # stat='identity' is used for summarized data.
  ggtitle("Attrition Maritial Status") +
  theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Overtime") +
  ylab("Count")

plot2 <- ggplot(df_summary2, aes(MaritalStatus, name_count, fill=MaritalStatus)) +  #
  geom_bar(stat = 'identity')+              # stat='identity' is used for summarized data.
  ggtitle("Non Attrition Maritial Status") +
  theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Overtime") +
  ylab("Count")

###################################################################################################
df_summary3 <-
  attrition_Employees                          %>% # Pipe df into group_by
  group_by(OverTime)              %>% # grouping by 'type' column
  summarise(name_count = n())     # calculate the name count for each group

df_summary4 <-
  no_attrition_Employees                          %>% # Pipe df into group_by
  group_by(OverTime)              %>% # grouping by 'type' column
  summarise(name_count = n())     # calculate the name count for each group


## (1) Plot pre summarized data: 'df_summary'.
plot3 <- ggplot(df_summary3, aes(OverTime, name_count, fill=OverTime)) +  #
  geom_bar(stat = 'identity')+              # stat='identity' is used for summarized data.
  ggtitle("Attrition Overtime") +
  theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Overtime") +
  ylab("Count")

plot4 <- ggplot(df_summary4, aes(OverTime, name_count, fill=OverTime)) +  #
  geom_bar(stat = 'identity')+              # stat='identity' is used for summarized data.
  ggtitle("Non Attrition Overtime") +
  theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Overtime") +
  ylab("Count")

grid.arrange(plot1, plot2, plot3, plot4, ncol=2, nrow=2)
```
## The Plots and T-Test to show inequality in montly income between Attrition and Non-Attrition Employees
```{r The Plots and T-Test to show inequality in montly income between Attrition and Non-Attrition Employees, echo=TRUE}
df_ddanalytics = read.csv("Source/CaseStudy2-data.csv", header = TRUE)
names(df_ddanalytics) <-c("Age","Attrition","BusinsTravl","DailyRate","Department","DistancHome","Education","EducField","EmployCount","EmployeeNum","EnvSatisfct","Gender","HourlyRate","JobInvolve","JobLevel","JobRole", "JobSatisfct","MaritlStats","MonthIncom","MonthRate", "NumCompWorkd","Over18", "OverTime","SalaryHike","PerfRating","RelSatisfct","StandHours","StockOption","WorkingYear","LastYrTrain","WorkLifeBal","YrsAtCompny","CurRoleYrs","YrslastProm","YrsCurMangr")
dfanalytics <- data.frame(df_ddanalytics$Age,df_ddanalytics$MonthIncom,df_ddanalytics$JobSatisfct,df_ddanalytics$DailyRate,df_ddanalytics$SalaryHike,df_ddanalytics$WorkLifeBal,df_ddanalytics$CurRoleYrs,df_ddanalytics$YrslastProm,df_ddanalytics$WorkingYear,df_ddanalytics$RelSatisfct, df_ddanalytics$Attrition)
##Summary of Attrition Employees
summary(df_ddanalytics %>% filter(Attrition =="Yes") %>% .$MonthIncom)

##Summary of Non-Attrition Employees
summary(df_ddanalytics %>% filter(Attrition =="No") %>% .$MonthIncom)

##Boxplot of Monthly Income Attrition vs Non-Attrition
ggplot(df_ddanalytics, aes(Attrition, MonthIncom, fill=Attrition)) +
  geom_boxplot() +
  ggtitle("Monthly Income") +
  theme(plot.title = element_text(hjust = 0.5))

##Histogram of Monthly Income Attrition vs Non-Attrition
p1 <- ggplot(df_ddanalytics, aes(MonthIncom)) +
  geom_histogram(fill = "blue", color = "grey30") +
  ggtitle("Plot of Employees Monthly Income") +
  theme(plot.title = element_text(hjust = 0.5)) +
  facet_wrap(~ Attrition)

##Histogram of Log(Monthly Income) Attrition vs Non-Attrition
p2 <- ggplot(df_ddanalytics, aes(MonthIncom)) +
  geom_histogram(fill = "red", color = "grey30") +
  ggtitle("Plot of Employees Log(Monthly Income)") +
  theme(plot.title = element_text(hjust = 0.5)) +
  facet_wrap(~ Attrition) +
  scale_x_log10()

##Plot the Histograms
grid.arrange(p1, p2, nrow = 2)

options(scipen=999) #disable scientific notation
##T-Test to test difference in the montly income means between Attrition vs Non-Attrition
t.test(MonthIncom ~ Attrition, data = df_ddanalytics)

##T-Test to test difference in the log(montly income) means between Attrition vs Non-Attrition
t.test(log(MonthIncom) ~ Attrition, data = df_ddanalytics)

##Non Parametric Test
wilcox.test(MonthIncom ~ Attrition, data = df_ddanalytics)


```